SUPPORTING INFORMATION:\\ Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-10-17 Manuscript/manuscriptPSsuppl.tex
Final edits on the SI.

@@

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela-Rosales}
\affiliation{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
\author{Tiago M. Ferreira}
\affiliation{NMR group - Institute for Physics, Martin-Luther University Halle-Wittenberg}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\alsoaffiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
and The Zernike Institute for Advanced Materials, 
University of Groningen, 9747 AG Groningen, The Netherlands}
\author{Markus Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jukka M{\"a}{\"a}tt{\"a}}
\affiliation{Department of Chemistry, Aalto University, Espoo, Finland}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas J. Piggot}
\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}

@@

@@

\caption{The list of POPC:POPS mixtures simulated with different amounts of added monovalent ions. 
  The salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
  This corresponds the concentration in buffer before solvating lipids, which are

@@

  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\

@@

\subsection{CHARMM36}

\noindent
    {\it POPC bilayer.} Previously published values \cite{botan15} calculated from the data
    available from Ref. \citenum{charmmPOPC512files} are used.
% It should be noted that pure POPC are ran with Gromacs 4, but the results with
% PS are ran with Gromacs 5. This has significant effect on acyl chains:
% https://github.com/NMRLipids/NmrLipidsCholXray/issues/4
% but the effect to headgroups is small
% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS4.5/Headgroup_Glycerol_Order_Parameters_Simulation.dat
% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS5.1.2/Headgroup_Glycerol_Order_Parameters_Simulation.dat}

\noindent
{\it DOPS and POPS bilayers.}
%The data in Table \ref{PSsystems} and Refs \citenum{charmm36DOPS303K,charmm36POPS298K,charmm36POPS298Kpotassium}
Starting structures for CHARMM36 DOPS and POPS simulations were constructed using the CHARMM-GUI webserver~\cite{lee16,jo18}.
The systems contained a total of 128 lipids (64 per leaflet of the membrane), 35 waters per lipid, and 128 Na$^+$ ions.
Force field parameters for DOPS and POPS were taken from Venable et al. and included the modified parameters for interactions
with Na$^+$ ions \cite{venable13}. Water was treated using the CHARMM TIP3P model \cite{durell94,eyal96}.
%https://pubs.acs.org/doi/10.1021/j100059a038 and https://aip.scitation.org/doi/10.1063/1.472061].
For POPS, an identical system was also constructed in which the 128 Na$^+$ ions were replaced with 128 K$^+$ ions.

Simulations of these systems were performed for 500~ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
A 2~fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
The simulations were maintained at temperatures of 303~K and 298~K for DOPS and POPS respectively using the Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}
with a coupling constant of 1~ps. A pressure of 1~bar was maintained using the Parrinello--Rahman \cite{parrinello81} method with a coupling constant of 5~ps.
Pressure coupling was applied in a semi-isotropic manner with the $x$ and $y$ dimensions, in the plane of the bilayer, fluctuating independently of the $z$ dimension.
Standard CHARMM36 force field methods were used for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
switched off between 1.0~nm and 1.2~nm; Coulombic interactions were truncated at 1.2~nm with long-range interactions treated using PME~\cite{darden93,essman95}. The final 100~ns of the simulations was used for analysis.

\noindent
{\it POPC:POPS mixtures without additional ions.}
Inputs for simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were configured using the CHARMM-GUI~\cite{lee16,jo18}. Two identical systems were constructed, apart from the counter ion; one system was neutralised through the addition of 22 Na$^+$ ions while the other system
contained 22 K$^+$ ions. As in the pure POPS CHARMM36 simulations, 35 water per lipid were added. All other system and simulation parameters
and settings were identical to those used in the CHARMM36 POPS simulations.

Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids (250:50 and 150:150, respectively)
with neutralizing potassium countertions 
%ran using CHARMM36 force field.
%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
%POPC:POPS (both 1:1 and 5:1) mixtures
were prepared
%with 300 lipids in total
%with the  ions
using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298~K. For further details see table \ref{mixedIONsystemsMONOVALENT}.

\noindent
{\it POPC:POPS (5:1) mixtures with additional monovalent ions.}
POPC:POPS (110:22) mixtures containing total 132 lipids with the additional
potassium or sodium ions corresponding concentrations of $\sim$450~mM and 890~mM
were generated with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using
Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation
parameters given by the CHARMM-GUI at 298~K. For further details, simulation files
and data, see table \ref{mixedIONsystemsMONOVALENT} and
Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}.

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium using NBfix1 parameters.} 
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
%were simulated with the additional calcium concentrations.
%The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI in January 2018 \cite{lee16,jo18}.
The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix1~\cite{kim16}.
Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
by the CHARMM-GUI at 298~K.

\noindent
{\it POPC:POPS (5:1) mixtures with  additional calcium using NBfix2 parameters.}
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids, neutralizing sodium ions, 
and 0.14~M or 0.94~M CaCl$_2$ were prepared using the CHARMM-GUI in August 2019 \cite{lee16,jo18}.
The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix2~\cite{han2018graph}. Systems were simulated using Gromacs 2018.6 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
by the CHARMM-GUI at 320~K.  \\

\subsection{CHARMM36ua}
\noindent
{\it DOPS and POPS bilayers.} 
%The data in Table \ref{PSsystems} and Refs \citenum{charmm36uaDOPS303K,charmm36uaPOPS298K}
Starting structures for CHARMM36-UA DOPS and POPS simulations were taken from those generated for the CHARMM36 simulations (see above),
with any extraneous hydrogen atoms removed in the lipid tails. Force field parameters for DOPS and POPS were constructed through a
combination of the published CHARMM36 PS~\cite{venable13} and CHARMM36-UA PC parameters~\cite{lee14}.
% [REF: https://pubs.acs.org/doi/10.1021/jp410344g].
All other simulation parameters and conditions were identical to those described for the CHARMM36 all-atom DOPS and POPS simulations.

\subsection{MacRog}
\noindent
{\it POPC bilayers.} 
% \todo{For example this data set has pretty good description of simulation details in the
% reference. I am not sure how much we should repeat these in here.}
The POPC bilayer consisting of a total of 128 lipids was constructed from single lipid structure taken 
from Ref.~\citenum{kulig15c}. The bilayer was hydrated by a total of 5120 water 
molecules, corresponding to 40 water molecules per lipid.
The MacRog lipid parameters \cite{kulig15b}, obtained from Ref.~\citenum{kulig15c}, were used,
along with the TIP3P water model~\cite{jorgensen83}. The simulation parameters were identical to those used for 
POPC/POPS mixtures with the MacRog force field, see below. The simulation was run for 200~ns using Gromacs 2016.3~\cite{abraham2015gromacs}, 
out of which 150~ns were used for analyses. The simulation data and related files are available from Ref. \citenum{macrogPOPC298K}.

\noindent
{\it POPS bilayers with Na$^+$ counterions.}
%The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
%\todo{T. Piggot and M. Javanainen, please write simulation details} \\
%\hl{Piggot: please write about system with Na$^+$ counterions}.
Force field parameters for POPS simulations with the MacRog OPLS-AA force field~\cite{maciejewski14,kulig15b,rog16}
were constructed manually using the PS parameters provided in ref.~\citenum{rog16} as a guide. The original parameters were
not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
The corrected parameters are available from Ref.~\citenum{macrogPOPS298KwithK}.
The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na$^+$ ions) and was converted into
the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
Simulation parameters were chosen to closely mimic those used in the original force field publications~\cite{maciejewski14,kulig15b,rog16}:
van der Waals interactions were truncated at 1.0~nm with a dispersion correction applied to the energy and pressure;
Long-range Coulombic interactions were treated with PME beyond a system-optimized cut-off.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
including use of the Verlet cut-off scheme~\cite{Pall13}, except that here LINCS was used to constrain all bonds.
Simulations with the group cut-off scheme gave slightly lower area per molecule ($\sim$~50\AA$^2$) than
the simulations with Verlet cut-off ($\sim$~52\AA$^2$).

\noindent
{\it POPS bilayers with K$^+$ counterions.}
The membrane with sodium counterions was further simulated after replacing Na$^+$ counterions by K$^+$ ones, which 
were described by the \AA{}qvist parameters \cite{aqvist90}. Other topologies
were unchanged. The simulation parameters were the same as used for the POPC/POPS mixtures with 
the MacRog force field, see below. The simulation was run for 200~ns, out of which 150~ns was used 
in the analyses.

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium ions.} 
%The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
two leaflets were set up using CHARMM-GUI \cite{lee16,jo18}.
The bilayers were solvated by 5760 water molecules, corresponding to 40 water molecules for lipid.
Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.
These bilayers were simulated using Gromacs 2016.3~\cite{abraham2015gromacs} in the presence of counterions only, as well as together with different
concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
were considered, which corresponded to the amounts of 10/20, 31/62, 104/208, and 311/622 
of Ca$^{2+}$/Cl$^-$ ions. For membranes with KCl, concentrations of 500~mM, 1~M, 2~M, and 3~M of 
were considered, corresponding to 52, 104, 208, or 311 pairs of K$^+$ and Cl$^-$ ions.

>>>>

%-- To remove the empty first page that comes from the too large title page:
\usepackage{atbegshi}% http://ctan.org/pkg/atbegshi
\AtBeginDocument{\AtBeginShipoutNext{\AtBeginShipoutDiscard}}

@@

\author{Hanne Antila}
\affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation[Moscow Institute of Physics and Technology]{Research Center for Molecular Mechanisms of Aging and Age-Related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny 141701, Russia}
\author{Fernando Favela-Rosales}
\affiliation[Tecnol\'{o}gico Nacional de M\'{e}xico]{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico Campus Zacatecas Occidente, C. P. 99102, Zacatecas, M\'{e}xico}
\author{Tiago M. Ferreira}
\affiliation[Martin-Luther University Halle--Wittenberg]{NMR Group - Institute for Physics, Martin-Luther University Halle--Wittenberg, 06120 Halle (Saale), Germany}
\author{Ivan Gushchin}
\affiliation{Research Center for Molecular Mechanisms of Aging and Age-Related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny 141701, Russia}
\author{Matti Javanainen}
\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the  Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, 60637 Chicago, Illinois, United States of America}
\affiliation[University of South Florida]{Department of Global Health, College of Public Health, University of South Florida, 33612 Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry of the Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\alsoaffiliation[University of Groningen]{Groningen Biomolecular Sciences and Biotechnology Institute and The Zernike Institute for Advanced Materials,  University of Groningen, 9747 AG Groningen, The Netherlands}
\author{Markus S. Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jukka M{\"a}{\"a}tt{\"a}}
\affiliation[Aalto University]{Department of Chemistry and Materials Science, Aalto University, 00076 Aalto, Finland}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry of the  Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry of the  Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\affiliation[University of Helsinki]{Institute of Biotechnology, University of Helsinki, 00014 Helsinki, Finland}
\author{Thomas J. Piggot}
\affiliation[University of Southampton]{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}

@@

\subsection{CHARMM36}
\label{sec:CHARMM36}
\noindent
    {\it POPC bilayer.} Previously published values \cite{botan15} calculated from the data
    available from Ref. \citenum{charmmPOPC512files} are used.
% It should be noted that pure POPC are ran with Gromacs 4, but the results with
% PS are ran with Gromacs 5. This has significant effect on acyl chains:
% https://github.com/NMRLipids/NmrLipidsCholXray/issues/4
% but the effect to headgroups is small
% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS4.5/Headgroup_Glycerol_Order_Parameters_Simulation.dat
% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS5.1.2/Headgroup_Glycerol_Order_Parameters_Simulation.dat}

\noindent
{\it DOPS and POPS bilayers.}
%The data in Table \ref{PSsystems} and Refs \citenum{charmm36DOPS303K,charmm36POPS298K,charmm36POPS298Kpotassium}
Starting structures %for CHARMM36 DOPS and POPS simulations
were constructed using the CHARMM-GUI webserver~\cite{lee16,jo18}.
The systems contained a total of 128 lipids (64 per leaflet of the membrane), 35 waters per lipid, and 128 Na$^+$ ions.
Force field parameters for DOPS and POPS were taken from Venable et al. and included the modified parameters for interactions
with Na$^+$ ions \cite{venable13}. Water was treated using the CHARMM TIP3P model \cite{durell94,eyal96}.
%https://pubs.acs.org/doi/10.1021/j100059a038 and https://aip.scitation.org/doi/10.1063/1.472061].
For POPS, an identical system was also constructed, in which the 128 Na$^+$ ions were replaced with 128 K$^+$ ions.

Simulations of these systems were performed for 500~ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
A 2~fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
The simulations were maintained at temperatures of 303~K for DOPS and 298~K for POPS using the Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}
with a coupling constant of 1~ps. A pressure of 1~bar was maintained using the Parrinello--Rahman \cite{parrinello81} method with a coupling constant of 5~ps.
Pressure coupling was applied in a semi-isotropic manner with the $x$ and $y$ dimensions, in the plane of the bilayer, fluctuating independently of the $z$ dimension.
Standard CHARMM36 force field methods were used for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
switched off between 1.0~nm and 1.2~nm; Coulombic interactions were truncated at 1.2~nm with long-range interactions treated using PME~\cite{darden93,essman95}. The final 100~ns of the simulations was used for analysis.

\noindent
{\it POPC:POPS mixtures without additional ions.}
Inputs for simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were created using the CHARMM-GUI~\cite{lee16,jo18}. Two identical systems were constructed, apart from the counter ion; one system was neutralised through the addition of 22 Na$^+$ ions while the other system
contained 22 K$^+$ ions. As in the pure POPS CHARMM36 simulations, 35 waters per lipid were added. All other system and simulation parameters
and settings were identical to those used in the CHARMM36 POPS simulations.

Simulations of POPC:POPS (5:1) and POPC:POPS (1:1) mixtures containing a total of 300 lipids (250:50 and 150:150)
with neutralizing potassium counterions 
%ran using CHARMM36 force field.
%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
%POPC:POPS (both 1:1 and 5:1) mixtures
were prepared
%with 300 lipids in total
%with the  ions
using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
and the CHARMM36 force field with simulation parameters given by the CHARMM-GUI at 298~K. For further details see Table \ref{mixedIONsystemsMONOVALENT}.

@@

\caption{POPC:POPS mixtures simulated with different amounts of added monovalent ions. 
  Salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
  This corresponds to the concentration in buffer before solvating lipids, as

@@

  lipids/counter-ion & force field for lipids / ion & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) &  \footnote{Number of lipid molecules}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) & \footnote{Reference for simulation files}files\\

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional monovalent ions.}
POPC:POPS (110:22) mixtures containing a total of 132 lipids, with additional
potassium or sodium ions corresponding to concentrations of $\sim$450~mM and 890~mM
were generated with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using
Gromacs 5 \cite{abraham2015gromacs} and the CHARMM36 force field with simulation
parameters given by the CHARMM-GUI at 298~K. For further details, simulation files
and data, see Table \ref{mixedIONsystemsMONOVALENT} and
Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}.

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium using the NBfix1 parameters.} 
POPC:POPS (5:1) mixtures containing a total of 300 (250:50) lipids
%were simulated with the additional calcium concentrations.
%The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI in January 2018 \cite{lee16,jo18}.
The ion parameters generated by the CHARMM-GUI at the time are labeled here as NBfix1~\cite{kim16}.
Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and the CHARMM36 force field with simulation parameters given
by the CHARMM-GUI at 298~K.

\noindent
{\it POPC:POPS (5:1) mixtures with  additional calcium using the NBfix2 parameters.}
POPC:POPS (5:1) mixtures containing a total of 300 (250:50) lipids, neutralizing sodium ions, 
and 0.14~M or 0.94~M CaCl$_2$ were prepared using the CHARMM-GUI in August 2019 \cite{lee16,jo18}.
The ion parameters generated by the CHARMM-GUI at the time are labeled here as NBfix2~\cite{han2018graph}. Systems were simulated using Gromacs 2018.6 \cite{abraham2015gromacs} and the CHARMM36 force field with simulation parameters given
by the CHARMM-GUI at 320~K.  \\

\subsection{CHARMM36-UA}
\noindent
{\it DOPS and POPS bilayers.} 
%The data in Table \ref{PSsystems} and Refs \citenum{charmm36uaDOPS303K,charmm36uaPOPS298K}
Starting structures %for CHARMM36-UA DOPS and POPS simulations
were taken from those generated for the CHARMM36 simulations (see section \ref{sec:CHARMM36}),
with any extraneous hydrogen atoms removed in the lipid tails. Force field parameters for DOPS and POPS were constructed through a
combination of the published CHARMM36 PS~\cite{venable13} and CHARMM36-UA PC parameters~\cite{lee14}.
% [REF: https://pubs.acs.org/doi/10.1021/jp410344g].
All other simulation parameters and conditions were identical to those described for the CHARMM36 all-atom DOPS and POPS simulations.

\subsection{MacRog}
\noindent
{\it POPC bilayers.} 
% \todo{For example this data set has pretty good description of simulation details in the
% reference. I am not sure how much we should repeat these in here.}
The POPC bilayer consisting of a total of 128 lipids was constructed from a single lipid structure taken 
from Ref.~\citenum{kulig15c}. The bilayer was hydrated by a total of 5120 water 
molecules, corresponding to 40 water molecules per lipid.
The MacRog lipid parameters \cite{kulig15b}, obtained from Ref.~\citenum{kulig15c}, were used,
along with the TIP3P water model~\cite{jorgensen83}. The simulation parameters were identical to those used for 
POPC:POPS (5:1) mixtures with the MacRog force field, see further below in this section. The simulation was run for 200~ns using Gromacs 2016.3~\cite{abraham2015gromacs}, 
out of which the last 150~ns were used for analysis. The simulation data and related files are available from Ref. \citenum{macrogPOPC298K}.

\noindent
{\it POPS bilayers with Na$^+$ counterions.}
%The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
%\todo{T. Piggot and M. Javanainen, please write simulation details} \\
%\hl{Piggot: please write about system with Na$^+$ counterions}.
Force field parameters for POPS simulations with the MacRog OPLS-AA force field~\cite{maciejewski14,kulig15b,rog16}
were constructed manually using the PS parameters provided in ref.~\citenum{rog16} as a guide. The original parameters were
not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
The corrected parameters are available from Ref.~\citenum{macrogPOPS298KwithK}.
The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na$^+$ ions), and was converted into
the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
Simulation parameters were chosen to closely mimic those used in the original force field publications~\cite{maciejewski14,kulig15b,rog16}:
van der Waals interactions were truncated at 1.0~nm with a dispersion correction applied to the energy and pressure;
long-range Coulombic interactions were treated with PME beyond a system-optimized cut-off.
All other simulation settings were identical to those described above in section \ref{sec:CHARMM36} for the CHARMM36 DOPS and POPS simulations,
including the use of the Verlet cut-off scheme~\cite{Pall13}, except that here LINCS was used to constrain all bonds.
Simulations with the group cut-off scheme gave slightly lower area per molecule ($\sim$~50\,\AA$^2$) than
the simulations with the Verlet cut-off ($\sim$~52\,\AA$^2$).

\noindent
{\it POPS bilayers with K$^+$ counterions.}
The membrane with sodium counterions was further simulated after replacing Na$^+$ counterions by K$^+$ ones, which 
were described by the \AA{}qvist parameters \cite{aqvist90}. Other topologies
were unchanged. Simulation parameters were the same as used for the POPC:POPS (5:1) mixtures with 
the MacRog force field, see further below in this section. The simulation was run for 200~ns, out of which the last 150~ns was used 
in the analysis.

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium ions.} 
%The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
Bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
two leaflets were set up using CHARMM-GUI \cite{lee16,jo18}.
The bilayers were solvated by 5760 water molecules, corresponding to 40 water molecules for lipid.
Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.
These bilayers were simulated using Gromacs 2016.3~\cite{abraham2015gromacs} in the presence of counterions only, as well as together with different
concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
were considered, which corresponded to the amounts of 10/20, 31/62, 104/208, and 311/622 
of Ca$^{2+}$/Cl$^-$ ions. For membranes with KCl, concentrations of 500~mM, 1~M, 2~M, and 3~M of 
were considered, corresponding to 52, 104, 208, and 311 pairs of K$^+$ and Cl$^-$ ions.

The lipids were described using the MacRog model \cite{maciejewski14,kulig15b,rog16}. The POPC topologies were obtained from 
Ref.~\citenum{kulig15c}, whereas the POPS topology was adapted from the OPPS topology in
Ref.~\citenum{rog16} by reversing the order of the acyl chains.
The initial structures from CHARMM-GUI were used, because they had the
naturally occurring L stereoisomer in the POPS head group, instead of the 
the D stereoisomer present in Ref.~\citenum{rog16}.
The TIP3P model \cite{jorgensen83} was used for water, and the \AA{}qvist parameters 
\cite{aqvist90}, standard for the OPLS force field, were employed for the ions.

Simulation parameters were taken from Ref.~\citenum{rog16}. 
Periodic boundary conditions were employed in all dimensions. The neighbor list was updated every step. 
%\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
\cite{essman95} was used for long-range electrostatics. Dispersion correction \cite{shirts07} 
was applied to both energy and pressure. Temperature was kept at 298~K using the 
Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}. The lipids and 
the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
\cite{parrinello81}. All non-water bonds were constrained with the LINCS algorithm 
\cite{hess97,hess07},
whereas the bonds in water molecules were constrained using SETTLE \cite{miyamoto92}.
The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 
(only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
250, 200, or 300~ns of the simulations were included in the analysis, respectively.

\subsection{Lipid17}
{\it DOPS and POPS bilayers.} 
The structures of the bilayers with 128 DOPS or 128 POPS lipids were
obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers were solvated with
4480 water molecules, resulting in 31.1 waters per lipid. For
each system, 128 Na$^+$ ions were added to neutralize the system.
In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. The SETTLE~\cite{miyamoto92} algorithm was employed 
to constrain the bonds in TIP3P water molecules. Simulations were run using the Amber18 simulation package,\cite{amber18md}
and lipid parameters were obtained from the Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with the Joung--Cheatham
ion parameters~\cite{joung2008determination} and another set of simulations with the ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
The data in Table~\ref{PSsystems} and Refs.~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
the number of molecules in simulations, related simulation files, and trajectories.
